When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-09 doc/paper/vpd_et_paper.tex
small typo fixed, send off to Pierre and Changjie for ocmments

% 11/23/2015
\documentclass[draft,linenumbers]{agujournal}
% \draftfalse
\drafttrue

@@

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved for later ET. An increase in ET in response to atmospheric drying will reduce soil moisture, but contribute increased moistening to the atmosphere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

Before diving into math and data, we have some prior intuition for how the system might behave. First, whether VPD drives or reduces ET should be a function of plant type. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Additionally, atmospheric conditions must matter as well. At the ecosystem scale there are limits to the strategies plants use to hold water back from the atmosphere. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

The purpose of this study is to examine ecosystem response to atmospheric drying during the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average GPP exceeds 10\% of the observed 95th percentile of GPP for a given site.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation and high relative humidity (\cite{MEDLYN2016).
\end{itemize}
Additionally, we restrict data to the daytime which is identified when downwelling shortwave radation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed. After these quality control proceedures, 332556 upscaled hourly observations remain.

@@

 where variable definitions are given in Table 1. \citet{MEDLYN_2011} developed a model for $g_s$ by combining optimal photosynthesis theory with empirical approaches. The result for leaf-scale stomatal resistance was:

@@

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so GPP in \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancelation of LAI in addition to uWUE in Equation \ref{medlyn}. Plugging Equation 5 into Equation 1 and rearranging gives:

@@

Given FLUXNET data described in Section \ref{data}, every term in Equation \ref{et} is known. However, our sampling of sites and our focus on the growing season may intoduce some deviations of $uWUE$ from those observed in \citet{Zhou_2015}. Also, we wish to include some measure of uncertainty in our analysis to guide if our many assumptions and simplifications are reasonable. To account for both mean deviations of $uWUE$ and uncertainty, we will introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

The variablility of $\sigma$ across sites and time will provides some measure of uncertainty in our model, assumptions, as well as the fluxnet observations themselves. To correct for differences in sampling between \cite{Zhou_2015} and our data, we set uWUE such that $\overline{\sigma} = 1$. The variability of $\sigma$ then propogates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

With Equation \ref{d_et} we have provided an analytical franework for ecosystem reponse to atmospheric demand perturbations. There are a few subtelties to taking the derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are functionally related, so while taking the derivative we evaluate $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD} \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH} \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$ fixed}}$\explain{Should I take the derivative this way? I think it's more mathematically correct, but it makes the explanation more complicated and only changes the solution by a factor of two.}. $RH$ and $e_s$ are assumed to be approximately orthogonal.

@@

One final comment on our derivation which will not be discussed further but is relevant for future analysis: if we  approximage $c_s$ at a global mean CO$_2$ concentration, then the RHS of Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015, Lin_2015}. This then begs the question, why use PET for drought metrics in vegetated areas? It appears a much more physically realistic estimate of ET can be had with the same information reqiured to calculate PET.

\begin{table}
\caption{Definition of symbols and variables, with citation for calculation if applicable.}
\centering
\begin{tabular}{l c c c}
\hline
 Variable & Description & Units & Citation \\
\hline
$e_s$  & saturation vapor pressure & Pa  & - \\ 
$T$  & temperature  & K & - \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
  $g_a$  & aerodynamic conductance & m s$^{-1}$  & \cite{Thom_1977}, \cite{Paulson_1970}, \cite{Beljaars_1991} \\
  $\rho_a$  & air density & kg m$^{-3}$  & - \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
  $VPD$  & vapor pressure deficit & Pa  & - \\
  $\gamma$  & psychrometric constant & Pa K$^{-1}$   & - \\
  $g_s$  & stomatal conductance & m s$^{-1}$  & \cite{MEDLYN_2011} \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  & \cite{MEDLYN_2011} \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
  $LAI$ & leaf area index & -& - \\
  $\sigma$ & uncertainty parameter & -& - \\
  $c_s$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

Before calculating the sensitivity of ET to VPD, we will consider the functional form of Equation \ref{d_et}. There are two main terms: a ``scaling'' term which modifies the mangnitude but not the sign:

@@

and a ``sign'' term which determines whether the derivative is postive or negative:

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will deterimine whether ET increases or decreases with increasing VPD. If the second term is larger magnitude then plant response dominates, while if the first term is larger then atmospheric demand dominats. Given that the sign term is crucial to answering our question ``When does VPD drive or reduce ET?'', we will explore its functional form first.

\subsection{Functional Form of the Sign Term}

The sign term is crucuial for answering our question ``When does VPD drive or reduce ET?,'' so let's explore the terms within it more. The variability of $c_s$ and $\gamma$ is relatively less than $\sigma$ and $VPD$ variability, so we sign term variability within PFT will approximately determined by just $\sigma$ and $VPD$. If we fix uncertainty $\sigma$ at 1, then the sign term is approximately just a function of $VPD$. We can further dertmine a $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

\begin{table}
  \label{vpd_crit}
\caption{Values of $VPD_{crit}$, where $\frac{\partial \; ET}{\partial \; VPD} = 0$, evaluated at PFT average values for $R_{air}$, $\sigma$, $\gamma$, and $c_s$. For reference, these values are also provided. For values of $VPD$ less than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be negative, and for values of $VPD$ greater than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be positive.}
\centering
\begin{tabular}{l c c c c c c}
  \hline
  PFT & $R_{air}$ & $c_s$ (ppm) & $\gamma$ & $\overline{\sigma}$ & $\overline{\sigma} \cdot uWUE$ & \textbf{$VPD_{crit}$ (Pa)} \\
  \hline
  CRO &  288.680920 & 372.567691& 65.351523& 0.684394&       2.602873&  \textbf{133.165438} \\
  CSH &   289.067152& 381.593622& 67.613172& 0.997224&       2.175278& \textbf{4439.564212} \\
  DBF &   288.624437& 377.449849& 63.421812& 0.881061&       2.746393&  \textbf{888.773243} \\
  ENF &  288.183849& 377.676463& 61.559242& 1.217892&       4.015362&  \textbf{978.084845} \\
  GRA &  288.425651& 377.264645& 61.598768& 0.850869&       2.281074& \textbf{1141.630778} \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

Figure \ref{term3} provides greater description for how the sign term varies with $VPD$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE$ and $g1$ shift the leading-order constant term ($\frac{1}{g_1}$) towards smaller values, and the sign term towards overall positive values. % Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water.\explain{really need to think about this sentence more, tehre is more nuance and I actually thing g1 and uwue really partially cancel.}
Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3, while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exert a greater influence than difference $uWUE$, as the power of $g1$ increases. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect due to its increasing powers. ENF ($g1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***need to replot with correct sigma (normalized to 1)***}}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of the minima of ET, as a function of $\sigma$ and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; VPD}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraction of positive $\frac{\partial \; ET}{\partial \; VPD}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; VPD}$ for CRO.

@@

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also impact statistics of $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variablility for overall $ET$ variability. So we will more closely examine the scaling term: $\frac{P}{T} \propto \rho$, so this should vary little relative to aerodynamic conductance and $\Delta$. $\gamma$ should also be relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature (through the function $\Delta$). This makes sense, as aerodynamic conductance represents how efficently response at the surface is communicated to the atmosphere. As it increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

$\Delta$'s presence in the scaling term also matches physical intition. Evaporative cooling will dampen the ability of the atmosphere to take more moisture, because $e_{s}$ decreases with decreasing temperature. The decrease in $e_{s}$ is proportional to $\Delta$ ($\delta e_{s} = \Delta \delta T$). So as $\Delta$ increases, you will get a larger damping of ET due to evaporative cooling.  The functional from of $\Delta$ will be the same across PFT, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness. So most of the differences in scaling between PFT should be in the aerodynamic conductance term. One interesting side note is that the coefficient of variability for both aerodynamic conductance and the scaling term is relatively constant across PFT, suggesting that the influence of aerodynamic conductance on the relative (to the PFT mean) variability of the scaling term is approximately similar across PFT.

Figure \ref{scale_vary}A shows scaling term normalized by mean aerodynamic conductance (calculated for each plant functional type), and confirms that much of the relative variability of the scaling term is caused by aerodynamic conductance variability. Generally, temperature (via $\Delta$) causes less relative variability. However, the impact of $T$ on the relative variability increases with increasing aerodynamic conductance.

While the relative variability of scaling term is similar across PFT, the absolute value of scaling term varies strongly across PFT. Figure \ref{scale_vary}B shows scaling term evaluated with the mean aerodynamic conductance for each PFT, and at the range of observed temperatures for each PFT. As expected, for the tree PFTs (DBF, ENF) the scaling term is much larger and the temperature dependence is much stronger. Systematic differences in observed temperatures also cause differences in the average magnitude of scaling term. For example, ENF experiences on average colder temperatures and is thus more likely to have a larger scaling term. Additionally, because the variability of aerodynamic conductance increases proportionally to the mean, the spread of the scaling term due to aerodynamic conductance variability will be larger for the tree PFTs, although this is not shown for simplicity. To summarize Figure \ref{scale_vary}, the variability of the scaling term within each PFT will look like Figure \ref{scale_vary}A for each PFT, but the scale of the y-axis will increase or decrease according to mean aerodynamic conductance oberved in Figure \ref{scale_vary}B.

@@

The units of $\frac{\partial \; ET}{\partial \; VPD}$ make it difficult to interpret if $VPD$ is even a meaningful contributor to ET's variability. To better understand $VPD$'s contribution, we mulitply $\frac{\partial \; ET}{\partial \; VPD}$ with $VPD$'s standard deviation to define a (linearized) relative change in ET for variations in $VPD$ . $VPD$'s contribution to ET's variability ranges between 30 - 40 W m$^{-2}$ for all PFTs except for CSH, which is about 100 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} \cdot std(R)$, as net radiation is generally the driver of ET (cite joe berry here)\explain{Find citation?}. For all PFTs except for CSH $VPD$ contributes between 30 - 40 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 88 \%. $VPD$'s variability is certiantly a non-negligable contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies with PFT. Large mean uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; VPD}$, but $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ is still negative for DBF due to its lower $g1$. ENF's low $g1$, even lower than DBF, increases the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ on $VPD$, and makes the function very nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and negative $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. Finally, the low $uWUE$ of CSH pushes it toward by far the lowest values $\frac{\partial \; ET}{\partial \; VPD}$ (Figure \ref{term3}). Variability in $VPD$ accounts for the largest about of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 30-40 \% of $R$'s contributions to ET variability.

>>>>

% 11/23/2015
\documentclass[draft,linenumbers]{agujournal}
% \draftfalse
\usepackage{makecell}
\drafttrue

@@

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of the land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved for later ET. An increase in ET in response to atmospheric drying will reduce soil moisture, but contribute increased moistening to the atmosphere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

Before diving into math and data, we have some prior intuition for how the system might behave. First, whether VPD drives or reduces ET should be a function of plant type. Plants that evolved to conserve water (e.g. arid shrubs) will be more likely to reduce ET with increasing VPD, and plants that have evolved to care little about water (e.g. crops) will be more likely to increase ET with increasing VPD. Atmospheric conditions must matter as well. At the ecosystem scale there are limits to the strategies plants use to hold water back from the atmosphere. As atmospheric demand for water (VPD) increases, ecosystems will begin to reach their water conservation limits. At this stage any further increase in VPD will most likely drive an increase in ET, because there is little more the plants can do keep water back.

@@

The purpose of this study is to examine ecosystem response to atmospheric drying during the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average GPP exceeds 10\% of the observed 95th percentile of GPP for a given site.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation and high relative humidity (\cite{MEDLYN_2011}).
\end{itemize}
Additionally, we restrict data to the daytime which is identified when downwelling shortwave radiation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed. After these quality control procedures, 332556 upscaled hourly observations remain.

@@

 where variable definitions are given in Table 1. \citet{MEDLYN_2011} developed a model for stomatal conductance ($g_s$) by combining optimal photosynthesis theory with empirical approaches. The result for leaf-scale stomatal resistance was:

@@

Note that $uWUE$ is fit on the ecosystem scale in \citet{Zhou_2015} so GPP in Equation \ref{uwue} is really $A\cdot \text{LAI}$. This leads to the cancelation of LAI in addition to uWUE in Equation \ref{medlyn}. Plugging Equation 5 into Equation 1 and rearranging gives:

@@

Given FLUXNET data described in Section \ref{data}, every term in Equation \ref{et} is known. However, our sampling of sites at the global scale may introduce some deviations of $uWUE$ from those observed in \citet{Zhou_2015}. Also, we wish to include some measure of uncertainty in our analysis to guide if our many assumptions and simplifications are reasonable. To account for both mean deviations of $uWUE$ and uncertainty, we will introduce an uncertainty parameter $\sigma$ modifying $uWUE$:

@@

The variability of $\sigma$ across sites and time will provides some measure of uncertainty in our model, assumptions, as well as the FLUXNET observations themselves. To correct for differences in sampling between \cite{Zhou_2015} and our data, we set uWUE such that $\overline{\sigma} = 1$. The variability of $\sigma$ then propagates through any uncertainty to our derivative of Equation \ref{et_sigma}:

@@

With Equation \ref{d_et} we have provided an analytical framework for ecosystem response to atmospheric demand perturbations.\explain{Should I take the derivative this way? I think it's more mathematically correct, but it makes the explanation more complicated and only changes the solution by a factor of two.} There are a few subtleties to taking the derivative in Equation \ref{d_et}: $\Delta$ ($\frac{d e_{s}}{d T}$) and $VPD$ are functionally related, so while taking the derivative we evaluate $\frac{\partial \; ET}{\partial \; VPD} = \frac{\partial \; ET} {\partial \; e_s} \frac{\partial \; e_s}{\partial \; VPD} \Big|_{\text{RH fixed}} + \frac{\partial \; ET}{\partial \; RH} \frac{\partial \; RH}{\partial \; VPD} \Big|_{\text{$e_s$ fixed}}$. $RH$ and $e_s$ are assumed to be approximately orthogonal.

@@

One final comment on our derivation which will not be discussed further but is relevant for future analysis: if we  approximate $c_s$ at a global mean CO$_2$ concentration, then the RHS of Equation \ref{et} is fully defined using commonly available weather station data and the constants published in \citet{Zhou_2015, Lin_2015}. This makes Equation \ref{et} a useful alternative to PET in drought indices and hydrometeorological analysis for vegetated surfaces. Equation \ref{et} better reflects the physics of water exchange at the land surface.

\begin{table}
\caption{Definition of symbols and variables, with citation for calculation if applicable.}
\centering
\begin{tabular}{l c c c}
\hline
 Variable & Description & Units & Citation \\
\hline
$e_s$  & saturation vapor pressure & Pa  & - \\ 
$T$  & temperature  & K & - \\
$\Delta$  & $\frac{\partial e_s}{\partial T}$ & Pa K$^{-1}$ & - \\
$R$  & net radiation at land surface minus ground heat flux & W m$^{-2}$   & - \\
  $g_a$  & aerodynamic conductance & m s$^{-1}$  & \makecell{\cite{Thom_1977} \\ \cite{Paulson_1970} \\ \cite{Beljaars_1991}} \\
  $\rho_a$  & air density & kg m$^{-3}$  & - \\
  $c_p$  & specific heat capacity of air at constant pressure & J K$^{-1}$ kg$^{-1}$ & - \\
  $VPD$  & vapor pressure deficit & Pa  & - \\
  $\gamma$  & psychometric constant & Pa K$^{-1}$   & - \\
  $g_s$  & stomatal conductance & m s$^{-1}$  & \cite{MEDLYN_2011} \\
  $g_{l-s}$  & leaf-scale stomatal conductance & mol m$^{-2}$ s$^{-1}$  & \cite{MEDLYN_2011} \\
  $R*$ & universal gas constant & J mol$^{-1}$ K$^{-1}$ & - \\
  $LAI$ & leaf area index & -& - \\
  $\sigma$ & uncertainty parameter & -& - \\
  $c_s$ & CO$_2$ concentration & $\mu$ mol CO$_2$ mol$^{-1}$ air& - \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

@@

Before calculating the sensitivity of ET to VPD, we will consider the functional form of Equation \ref{d_et}. There are two main terms: a ``scaling'' term which modifies the magnitude but not the sign:

@@

and a ``sign'' term which determines whether the derivative is positive or negative:

\begin{equation}
  \label{sign}
  \frac{c_p}{R_{air}} - \frac{\gamma c_s }{1.6 \; R\; \text{ uWUE }} \left( \frac{2 g_1 + \sqrt{VPD}}{2 (g_1 + \sqrt{VPD})^2}\right).
\end{equation}
All variables are positive, so the relative magnitude between the first term and the second term in Equation \ref{sign} will determine whether ET increases or decreases with increasing VPD. If the second term is larger magnitude then plant response dominates, while if the first term is larger then atmospheric demand dominates. Given that the sign term is crucial to answering our question ``When does VPD drive or reduce ET?'', we will explore its functional form first.

\subsection{Functional Form of the Sign Term}
\label{sign_func}
First, let's explore the variables within the sign term to gain better intuition for what determines when VPD drives vs. reduces ET. The variability of $c_s$ and $\gamma$ is relatively less than $\sigma$ and $VPD$ variability, so sign term variability within PFT will be approximately determined by just $\sigma$ and $VPD$. If we fix uncertainty $\sigma$ at 1, then the sign term is approximately just a function of $VPD$. We can further determine a $VPD_{crit}$ where $\frac{\partial \; ET}{\partial \; VPD} = 0$:

@@

\begin{table}
  \label{vpd_crit}
\caption{Values of $VPD_{crit}$, where $\frac{\partial \; ET}{\partial \; VPD} = 0$, evaluated at PFT average values for $R_{air}$, $\sigma$, $\gamma$, and $c_s$. For reference, these values are also provided. For values of $VPD$ less than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be negative, and for values of $VPD$ greater than $VPD_{crit}$, $\frac{\partial \; ET}{\partial \; VPD}$ will be positive. \textbf{**** this statistics are dated, need to update***}}
\centering
\begin{tabular}{l c c c c c c}
  \hline
  PFT & $R_{air}$ & $c_s$ (ppm) & $\gamma$ &  uWUE    & \textbf{$VPD_{crit}$ (Pa)} \\
  \hline
  CRO &  288.680920 & 372.567691& 65.351523& 2.602873&  \textbf{133.165438} \\
  CSH &   289.067152& 381.593622& 67.613172& 2.175278& \textbf{4439.564212} \\
  DBF &   288.624437& 377.449849& 63.421812& 2.746393&  \textbf{888.773243} \\
  ENF &  288.183849& 377.676463& 61.559242& 4.015362&  \textbf{978.084845} \\
  GRA &  288.425651& 377.264645& 61.598768& 2.281074& \textbf{1141.630778} \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

Figure \ref{term3} provides a greater description for how the sign term varies with $VPD$, as a function of PFT. Equation \ref{d_et_approx} aids interpretation of Figure \ref{term3}. Larger $uWUE$ and $g1$ shift the leading-order constant term ($\frac{1}{g_1}$) towards smaller values, and the sign term towards overall positive values. % Higher values corresponde to plants that are more willing to spend water on primary production and less evolved to conserve water.\explain{really need to think about this sentence more, tehre is more nuance and I actually thing g1 and uwue really partially cancel.}
Figure \ref{term3} confirms our physical intuition: CROs are the least water conservative so have the smallest constant portion of Term 3 ($\frac{\gamma c_s }{1.6 \; R*\; \sigma \; \text{ uWUE }} \left(\frac{1}{g_1}\right)$), while CSH are the most water conservative and have the largest constant portion of Term 3. For the VPD-dependent terms in Equation \ref{d_et_approx}, differences in $g1$ between PFTs exert a greater influence than differences in $uWUE$. Increasing $uWUE$ and $g1$ decreases the VPD-dependence, but $g1$ has the bigger effect, particularly at high VPD when the series becomes disordered. ENF ($g1 = 74.31$) has by far the largest VPD dependence of response, while CRO ($g1 = 183.1$) has the smallest VPD dependence.

A primary takeaway from Figure \ref{term3}a is that according to our theory for all PFTs except for crops there is frequent occurrence of a negative (plant dominating) ET response to increases in VPD. To better illustrate this, the ranges of observed environmental VPDs at the FLUXNET sites are plotted parallel to the x-axis. 25th, 50th, and 75th percentiles are marked by stars, and the location of VPD$_{crit}$ is marked by a vertical line. For CSH (water conservative), VPD is always less than VPD$_{crit}$ (plant response dominates). For CRO, VPD is almost always more than VPD$_{crit}$. For forests and grass, whether plant response or atmospheric demand dominates will depend more on the environment, with about half of of observed VPD less than VPD$_{crit}$ (plant response dominates). It's also important to note that even when atmospheric demand dominates, the response is still far smaller than it would be for PET (atmospheric demand only). The sign term in this case would just be a constant ($\frac{c_p}{R_{air}} \approx 3.5$), which is far larger than any part of the curves for any PFT. This highlights the deficiencies of PET's ability to capture land response to changes in atmospheric dryness (VPD). However, up to this point we have assumed $\sigma$ (uncertainty) equal to one. We still need to test if inclusion of uncertainty changes the story.
\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig05.pdf}
\caption{Sources of variability for the sign term. Top: sign term as a function of VPD, with $\sigma$ held fixed at 1. The observed range of VPD for each PFT is also shown below the x-axis. Line extent corresponds to 5th and 95th percentiles, while stars denote the location of the 25th, 50th, and 75th percentiles.

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***need to replot with correct sigma (normalized to 1), also correct uWUE label in first plot***}}
\label{term3}
\end{figure}

Figure \ref{term3}b shows the location of VPD$_{crit}$, as a function of $\sigma$ (uncertainty) and $VPD$. For any $\sigma$ or VPD less (more) than these curves, the sign term will be negative (positive). It is clear that the portion of VPD observations below/above these curves will be a strong function of $\sigma$. However, we can see some general trends. For CSH, $\frac{\partial \; ET}{\partial \; VPD}$ should be negative for the vast majority of observed $\sigma$ and VPD. The fraction of positive $\frac{\partial \; ET}{\partial \; VPD}$ appears to be more even for ENF, GRA, and DBF, and we might expect a greater frequency of positive $\frac{\partial \; ET}{\partial \; VPD}$ for CRO. Section \ref{testing} further examines the consistency of our theory, given  the apparent sensitivity to uncertainty ($\sigma$).

@@

While the above discussion of the sign of $\frac{\partial \; ET}{\partial \; VPD}$ is important to answer our research question, the magnitude of $\frac{\partial \; ET}{\partial \; VPD}$ will also impact statistics of $\frac{\partial \; ET}{\partial \; VPD}$ and the importance of $VPD$ variability for overall $ET$ variability. So we will more closely examine the scaling term: $\frac{P}{T} \propto \rho$, so this should vary little relative to aerodynamic conductance and $\Delta$. $\gamma$ should also be relatively constant, so the scaling term should be primarily a function of aerodynamic conductance and temperature (through the function $\Delta$). This makes sense, as aerodynamic conductance represents how efficiently response at the surface is communicated to the atmosphere. As it increases, any plant response will be communicated more strongly to the atmosphere (and vice-versa).

$\Delta$'s presence in the scaling term also matches physical intuition. Evaporative cooling will dampen the ability of the atmosphere to take more moisture, because $e_{s}$ decreases with decreasing temperature. The decrease in $e_{s}$ is proportional to $\Delta$ ($\delta e_{s} = \Delta \delta T$). So as $\Delta$ increases, you will get a larger damping of ET due to evaporative cooling.  The functional from of $\Delta$ will be the same across PFT, but the temperature range may vary slightly. In contrast, aerodynamic conductance will vary strongly with PFT due to the importance of surface roughness. So most of the differences in scaling between PFT should be in the aerodynamic conductance term. One interesting side note is that the coefficient of variability for both aerodynamic conductance and the scaling term is relatively constant across PFT, suggesting that the influence of aerodynamic conductance on the relative (to the PFT mean) variability of the scaling term is approximately similar across PFT.

Figure \ref{scale_vary}A shows the scaling term normalized by mean aerodynamic conductance (calculated for each plant functional type), and confirms that much of the relative variability of the scaling term is caused by aerodynamic conductance variability. Generally, temperature (via $\Delta$) causes less relative variability. However, the impact of $T$ on the relative variability increases with increasing aerodynamic conductance.

While the relative variability of scaling term is similar across PFT, the absolute value of scaling term varies strongly across PFT. Figure \ref{scale_vary}B shows scaling term evaluated at the mean aerodynamic conductance for each PFT, and at the range of observed temperatures for each PFT. As expected, for the tree PFTs (DBF, ENF) the scaling term is much larger and the temperature dependence is much stronger. Systematic differences in observed temperatures also cause differences in the average magnitude of scaling term. For example, ENF experiences on average colder temperatures and is thus more likely to have a larger scaling term. Additionally, because the variability of aerodynamic conductance increases proportionally to the mean, the spread of the scaling term due to aerodynamic conductance variability will be larger for the tree PFTs, although this is not shown for simplicity. To summarize Figure \ref{scale_vary}: the variability of the scaling term within each PFT will look like Figure \ref{scale_vary}A for each PFT, but the scale of the y-axis will increase or decrease according to mean aerodynamic conductance observed in Figure \ref{scale_vary}B.

@@

The units of $\frac{\partial \; ET}{\partial \; VPD}$ make it difficult to interpret if $VPD$ is even a meaningful contributor to ET's variability. To better understand $VPD$'s contribution, we multiply $\frac{\partial \; ET}{\partial \; VPD}\left(\overline{env}\right)$ with $VPD$'s standard deviation to define a (linearized) relative change in ET for variations in $VPD$ . $VPD$'s contribution to ET's variability ranges between 30 - 40 W m$^{-2}$ for all PFTs except for CSH, which is about 100 W m$^{-2}$. Another meaningful comparison is to $\frac{\partial \; ET}{\partial \; R} \cdot std(R)$, as net radiation is generally the driver of ET (cite joe berry here)\explain{Find citation?}. For all PFTs except for CSH $VPD$ contributes between 30 - 40 \% of $R$'s contribution to variability. For CSH the portion is much larger, about 88 \%. $VPD$'s variability is certainly a non-negligible contributor to $ET$'s variability.

So far, idealized plots and statistics have illuminated the form of $\frac{\partial \; ET}{\partial \; VPD}$ and how it varies with PFT. Large mean uWUE shifts CRO and DBF towards positive $\frac{\partial \; ET}{\partial \; VPD}$, but $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ is still negative for DBF due to its lower $g1$. ENF's low $g1$, even lower than DBF, increases the dependence of $\frac{\partial \; ET}{\partial \; VPD}$ on $VPD$, and makes the function very nonlinear. This has the side effect of pushing $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ negative further than other PFTs for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and magnitude $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. GRA shows the opposite behavior; a relatively high $g1$  makes the function more linear, decreasing the magnitude of $\overline{\frac{\partial \; ET}{\partial \; VPD}}$ for a given fraction $\frac{\partial \; ET}{\partial \; VPD} < 0$ and negative $\frac{\partial \; ET}{\partial \; VPD}(\overline{T,\ldots,VPD})$. Finally, the low $uWUE$ of CSH pushes it toward by far the lowest values $\frac{\partial \; ET}{\partial \; VPD}$ (Figure \ref{term3}). Variability in $VPD$ accounts for the largest amount of $ET$ variability for CSH. For the other PFTs, $VPD$ contributes less to $ET$ variability, but still represents about 30-40 \% of $R$'s contributions to ET variability.
